---
title: "Въведение в анализ на времеви редове"
author: "Boyko Amarov"
date: "10/12/2021"
output: html_document
---

# Time series in R

```{r}
library(xts)
library(ggplot2)

bitcoins <- read.csv(
  'https://s3.eu-central-1.amazonaws.com/sf-timeseries/data/bitcoin2017-06-01-2017-11-13.csv', 
  stringsAsFactors = FALSE
)
bc_ts <- xts(bitcoins$close, order.by = as.Date(bitcoins$Index))
```


```{r}
plot(bc_ts)
```

$$
y_{1}, y_{2}, y_{3}, \ldots, y_{T}\\
y_{t}, t = 1,\ldots, T
$$
In the bitcoins example $T = 117$. $y_1 = 2475.248, y_2 = 2688.172, \ldots, y_{117} = 6535.948$.

Looking from Nov 13th 2017: what was the value on the previous trading day? Answer
by simply looking in the table: $y_{117 - 1} = y_{116} = 6622.517$. 

$$
y_{t}, t = 1,\ldots, T\\
y_{t - 1}: \text{ first lag of } y_{t}\\
y_{t - 2}: \text{ second lag of } y_{t}
$$

```{r}
bc_ts_l1 <- lag(bc_ts)
bc_ts_l2 <- lag(bc_ts, k = 2)

ggplot(data = bc_ts_all_lags, aes(x = bc_ts_l1, y = bc_ts)) +
  geom_point() +
  labs(
    title = "Scatterplot of y[t] and y[t-1]",
    x = expression(y[t-1]),
    y = expression(y[t])
  )
```

```{r}
mean(bc_ts_l1, na.rm = TRUE)
```

```{r}
rnorm(n = 1, mean = 10)
```

```{r}
# Изтегляме сто стойности от стандартно нормално разпределение

xrandom <- rnorm(n = 100)
ts_random <- xts(xrandom, order.by = seq.Date(from = as.Date("2017-06-01"), by = "day", length.out = 100))
```

```{r}
plot(ts_random)
```

```{r}
ts_random_l1 <- lag(ts_random)
ts_all_lags <- cbind(ts_random, ts_random_l1)

ggplot(ts_all_lags, aes(x = ts_random_l1, y = ts_random)) + 
  geom_point() + 
  ggtitle('') +
  ylab('y[t]') + 
  xlab('y[t - 1]')
```


$$
y_1, y_2, \ldots, y_T
$$
## World of mathematics (properties of baskets)

Mathematical expectation (expected value, mean)
$$
E(y_t)
$$
this will describe the average (in some sense) of the numbers in the basket.

$$
Var(y_t)
$$
Variance: this will describe how different the number in the basket are.

$$
Cov(y_t, y_{t - 1})
$$
Covariance: this will describe how the different baskets are related.

## World of observations

1. Empirical (sample) average (mean)

$$
\bar{y} = \frac{1}{T}\sum_{t = 1}^{T} y_{t}
$$

2. Empirical (sample) variance

$$
S^2_{y} = \frac{1}{T}\sum_{t = 1}^{T}(y_t - \bar{y})^2
$$

Exercise:

$$
y_1 = 2, y_2 = 3, y_3 = 10
$$

$$
\bar{y} = \frac{2 + 3 + 10}{3} = 5
$$
$$
S^2_{y} = \frac{1}{3}\left((y_1 - \bar{y})^2 + (y_2 - \bar{y})^2 + (y_3 - \bar{y})^2 \right)=\\
S^2_{y} = \frac{1}{3}\left((2 - 5)^2 + (3 - 5)^2 + (10 - 5)^2 \right)=\\
S^2_{y} = \frac{1}{3}\left(9 + 4 + 25\right) \approx 12.6
$$

```{r}
y <- c(2, 3, 10)
mean(y)
var(y) # Up to the scaling factor
```





3. Empirical (sample) covariance

$$
S^2_{y_t, y_{t-1}} = \frac{1}{T}\sum_{t = 1}^{T}(y_t - \bar{y})(y_{t - 1} - \bar{y})
$$






