---
title: "DF-test"
author: "Boyko Amarov"
date: "1/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(urca)
```


Random walk model without a drift

$$
y_{t} = y_{t - 1} + u_{t}, u_t \sim WN(\sigma^2)
$$

$$
\lambda - 1 = 0 \implies 
\lambda^* = 1
$$

The solution of this characteristic equation is 1. We say that the characteristic equation has a unit root.

## Random walks simulation 

```{r}
rw1 <- arima.sim(model = list(order = c(0, 1, 0)), n = 200)
rw2 <- arima.sim(model = list(order = c(0, 1, 0)), n = 200)

# rw2 <- 1.5 * rw1 + rnorm(n = 200)
plot(rw1)
plot(rw2)
```

```{r}
summary(lm(rw2 ~ rw1))
```
Regression model
$$
y_{t}: \text{ random walk 2}\\
x_{t}: \text{ random walk 1}\\
y_{t} = \beta_0 + \beta_1 x_{t} + e_{t}, e_{t} \sim N(0, \sigma^2)
$$
Estimated regression equation

$$
\hat{y}_{t} = 6.15 - 1.78 x_{t}
$$
T-Test of the hypothesis 

$$
H_0: \beta_1 = 0\\
H_1: \beta_1 \neq 0
$$
We reject the null hypothesis if the p-value of the test is less than 0.05 (convention).

spurious regression



$$
\hat{y}_{t} = -1.86 + 0.19 x_{t}
$$


# Unit root tests

AR(1) process without a constant

$$
y_{t} = \phi_1 y_{t - 1} + u_{t}, u_t \sim WN(\sigma^2)
$$
If the process has a unit root, it follows that $\phi_1 = 1$.

We would like to test the hypothesis:

$$
H_0: \phi_1 = 1\\
H_1: \phi < 1
$$
First difference
$$
\Delta y_{t} = y_{t} - y_{t - 1}
$$

First difference of a random walk

$$
y_{t} = y_{t - 1} + u_{t}\\
y_{t} - y_{t - 1} = u_{t}\\
\Delta y_{t} = u_{t}
$$

The first difference of the random walk process is stationary, because it is 
simply equal to the pure random process $u_{t}$ and it is stationary by the properties of the pure random process. We say that the process (in this case the random walk process) is integrated of order 1.

